A Fast kNN Algorithm Using Multiple Space-Filling Curves

### 2022-04-11 2022/Sensors/sensors.tex
add new fig files with .pdf extension
fix some typos

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{
The paper considers a time-efficient implementation of the k-nearest neighborhood (kNN) algorithm. A well-known approach for accelerating the kNN algorithm is to utilize dimensionality reduction methods based on the use of space-filling curves. In this paper, we take this approach further and propose an algorithm that employs multiple space-filling curves and is faster (with comparable quality) compared to the kNN algorithm which uses kd-trees to determine the nearest neighbors. A specific method for constructing multiple Peano curves is outlined, and statements are given about the preservation of object proximity information in the course of dimensionality reduction. An experimental comparison with known kNN implementations using kd-trees was performed using test and real-life data.
}

@@

In many cases, solving application problems comes down to the problem of classification, i.e. assigning the object under study to one of the available classes. One of the well-known classification methods is the kNN (k-nearest neighbourhood) method. Although the earliest papers related to this method appeared more than $50$ years ago \cite{Cover1967, Cover1968}, theoretical studies of various aspects of using kNN are still going on \cite{Pestov2013,Mirkes2020}. This is not to mention hundreds of publications in which this method is used to solve applied classification problems. Over many years of use, kNN has established itself as a simple and reliable method that yields acceptable results when solving many different problems.

@@

The paper is organized as follows. Section 2 describes the scheme for constructing a Peano-Hilbert curve approximation (evolvent) and the kNN method, which uses the evolvent to reduce the dimensionality of the data. The fundamental drawbacks of this dimensionality reduction method are noted. Section 3 proposes a way to overcome the drawbacks mentioned, based on the use of multiple evolvents. A theoretical statement is formulated about preserving some proximity information in the multidimensional space on one of the one-dimensional scales. A kNN scheme using multiple evolvents is presented. Section 4 contains experimental results comparing different implementations of kNN using synthetic and real-life data. Section 5 concludes the paper.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Problem Statement}

In this study, the classification problem will be understood as the problem of assigning an object $\omega$  to one of the predetermined classes based on its features. We will assume that each of the objects $\omega$ is represented as a feature vector 
\begin{equation}
y = (y_1, y_2, ..., y_N) \in R^N,
\end{equation} 
where the value of the $j$-th vector coordinate corresponds to the numerical value of the $j$-th feature of the object $\omega$.

We will also assume that there is already some number of objects exactly classified (a test sample), i.e. for each object we know which class it belongs to.

The $k$ nearest neighbours (kNN) method is based on the following simple rule: an object is considered to belong to the class to which most of its nearest neighbours belong. Here, "neighbours" refers to objects that are close (in one sense or another) to the one being studied.
Note that in order to apply this method, some metric $L(\omega_i, \omega_j)$, i.e. the distance function, should be introduced in the feature space of objects. As a rule, the Euclidean distance is employed here, although other metrics may also be used \cite{Mirkes2020}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Dimensionality reduction using space-filling curves}

@@

When $K$ neighbours need to be found, this search procedure can be repeated $K$ times. In this case it is necessary to exclude previously found items from the search. This can be done by excluding $x_i$ from the set being searched or by comparing the found $x_i$ with the previously found ones to find a match. For example, one possible implementation of searching for $K$ nearest neighbours could be as follows: in relation to $x^*$, we look for the nearest values to the left and right on the one-dimensional scale, choose the minimum one and continue searching sideways until exactly K neighbours are found.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Constructing a family of Peano curves}

@@

The scheme of implementation of the kNN algorithm using multiple Peano curves will look as follows.

For every object $\omega_i$ a set of its preimages on different one-dimensional scales is assigned 
\begin{equation}
\omega_i \rightarrow \{ x_i^0, x_i^1, ..., x_i^L \}
\end{equation} 
where $l, \; 0 \leq l \leq L$, is the number of the one-dimensional interval.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Special features of the implementation of the kNN method using multiple space-filling curves}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Experimental Results}

@@

The study was performed using several datasets, both randomly generated for testing and real-world  datasets from the database [https://archive.ics.uci.edu/ml/index.php]:
\begin{enumerate}
\item
\textbf{DS-Random-1} -- a set of random data generated using the random module of the Python NumPy library. The data has a uniform distribution and contains values on an interval from 0 to 1.

@@

Two kinds of studies were carried out:
\begin{itemize}
\item The time required to find K nearest neighbours for a test object was measured;
\item The accuracy of recognition of test objects was studied, based on the analysis as to whether their neighbours belong to certain classes.
\end{itemize}

@@

\begin{figure}
\widefigure
\includegraphics[width=1.0\linewidth]{fig3.png}
\caption{Running times of the algorithms with $N = 2$\label{fig3}}
\end{figure}

@@

\begin{figure}
\widefigure
\includegraphics[width=1.0\linewidth]{fig4.png}
\caption{Running times of the algorithms with $N = 3$\label{fig4}}
\end{figure}

\begin{figure}
\widefigure
\includegraphics[width=1.0\linewidth]{fig5.png}
\caption{Running times of the algorithms with $N = 4$\label{fig5}}
\end{figure}

@@

\begin{figure}[h]
\widefigure
\includegraphics[width=0.5\linewidth]{fig6.png}
\caption{Comparison of neighbour search times when determining the colour of an image point\label{fig6}}
\end{figure}

@@

The time required to find K nearest neighbours on the CarEvoluation dataset is shown in Fig.~\ref{fig6a}. In this experiment, the number of evolvents used simultaneously was 7. 
\begin{figure}
\widefigure
\includegraphics[width=0.5\linewidth]{fig6a.png}
\caption{Comparison of neighbour search times for the CarEvoluation dataset\label{fig6a}}
\end{figure}
The results of the comparison on this dataset also demonstrate that the kNN-ME method is faster compared to the kNN-KD method.

@@

\begin{specialtable}[H] 
  \caption{Percentage of errors in DS-Random-2 recognition}\label{tab:t3}
	\center
	\begin{tabular}{ccccccccc}
\toprule
K & \multicolumn{2}{c}{ $N=2$ } & & \multicolumn{2}{c}{$N=3$} & & \multicolumn{2}{c}{$N=4$} \\
		\noalign{\smallskip} \cline{2-3} \cline{5-6} \cline{8-9} \noalign{\smallskip}
  & kNN-KD & kNN-ME & & kNN-KD & kNN-ME & & kNN-KD & kNN-ME \\
\midrule													
5  & 5 & 5 & & 11 & 12 & & 6 & 8 \\
15 & 4 & 6 & & 10 & 11 & & 10 & 9 \\
25 & 5 & 7 & & 10 & 11 & & 7 & 9 \\
35 & 6 & 6 & & 11 & 13 & & 7 & 10 \\
45 & 7 & 6 & & 10 & 13 & & 7 & 10 \\
\bottomrule
\end{tabular}
\end{specialtable}

@@

The results of measuring the recognition accuracy on the real-world DS-Skin dataset show that on real-world data both methods show the same percentage of misclassification. The specific error percentages obtained with different numbers of neighbours K are shown in Table \ref{tab:t4}.

\begin{specialtable}[H] 
  \caption{Percentage of DS-Skin recognition errors, $N=3$}\label{tab:t4}
	\center
\begin{tabular}{ccc}
\toprule
K & kNN-KD & kNN-ME \\
\midrule													
5 & 0.5 & 0.5 \\
15 & 0.5 & 0.5 \\
25 & 0.5 & 0.5\\
35 & 1 & 1\\
45 & 1 & 1\\
\bottomrule
\end{tabular}
\end{specialtable}

@@

\begin{specialtable}[H] 
  \caption{Average error rate (in percent) for DS-CarEvaluation recognition, $N = 6$}\label{tab:t5}
	\center
\begin{tabular}{ccc}
\toprule
K & kNN-KD & kNN-ME \\
\midrule													
5 & 8 & 11 \\
7 & 9 & 10 \\
9 & 8 & 10 \\
11 & 10 & 11 \\
13 & 9 & 11 \\
\bottomrule
\end{tabular}
\end{specialtable}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}

The paper considers a classical machine learning method, the k nearest neighbour (kNN) method, which is widely used in machine learning practice and has been proven to perform very well because of its ``explainability'' (the decision to select an object class can be clearly explained). The results of practical implementation of one of kNN acceleration approaches based on dimensionality reduction using space-filling curves are presented.

@@

\dataavailability{The data analyzed in this study are openly available at https:// .}

>>>>

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{
The paper considers a time-efficient implementation of the $k$ nearest neighbours (kNN) algorithm. A well-known approach for accelerating the kNN algorithm is to utilize dimensionality reduction methods based on the use of space-filling curves. In this paper, we take this approach further and propose an algorithm that employs multiple space-filling curves and is faster (with comparable quality) compared to the kNN algorithm which uses kd-trees to determine the nearest neighbors. A specific method for constructing multiple Peano curves is outlined, and statements are given about the preservation of object proximity information in the course of dimensionality reduction. An experimental comparison with known kNN implementations using kd-trees was performed using test and real-life data.
}

@@

In many cases, solving application problems comes down to the problem of classification, i.e. assigning the object under study to one of the available classes. One of the well-known classification methods is the kNN ($k$ nearest neighbours) method. Although the earliest papers related to this method appeared more than $50$ years ago \cite{Cover1967, Cover1968}, theoretical studies of various aspects of using kNN are still going on \cite{Pestov2013,Mirkes2020}. This is not to mention hundreds of publications in which this method is used to solve applied classification problems. Over many years of use, kNN has established itself as a simple and reliable method that yields acceptable results when solving many different problems.

@@

The paper is organized as follows. Section \ref{ps} contains the problem statement and description of the kNN method. Section \ref{Peano} describes the scheme for constructing a Peano-Hilbert curve approximation (evolvent) and the kNN method, which uses the evolvent to reduce the dimensionality of the data. The fundamental drawbacks of this dimensionality reduction method are noted. Section \ref{MPeano} proposes a way to overcome the drawbacks mentioned, based on the use of multiple evolvents. A theoretical statement is formulated about preserving some proximity information in the multidimensional space on one of the one-dimensional scales. A kNN scheme using multiple evolvents is presented in Section \ref{knnme}. Section \ref{results} contains experimental results comparing different implementations of kNN using synthetic and real-life data. Section \ref{conclusions} concludes the paper.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Problem Statement}\label{ps}

In this study, the classification problem will be understood as the problem of assigning an object $\omega$  to one of the predetermined classes based on its features. We will assume that each of the objects $\omega$ is represented as a feature vector 
\begin{equation}
y = (y_1, y_2, ..., y_N) \in R^N,
\end{equation} 
where the value of the $j$-th vector coordinate corresponds to the numerical value of the $j$-th feature of the object $\omega$.
We will also assume that there is already some number of objects exactly classified (a test sample), i.e. for each object we know which class it belongs to.

The $k$ nearest neighbours (kNN) method is based on the following simple rule: an object is considered to belong to the class to which most of its nearest neighbours belong. Here, ``neighbours'' refers to objects that are close (in one sense or another) to the one being studied.
Note that in order to apply this method, some metric $L(\omega_i, \omega_j)$, i.e. the distance function, should be introduced in the feature space of objects. As a rule, the Euclidean distance is employed here, although other metrics may also be used \cite{Mirkes2020}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Dimensionality reduction using space-filling curves}\label{Peano}

@@

When $K$ neighbours need to be found, this search procedure can be repeated $K$ times. In this case it is necessary to exclude previously found items from the search. This can be done by excluding $x_i$ from the set being searched or by comparing the found $x_i$ with the previously found ones to find a match. For example, one possible implementation of searching for $K$ nearest neighbours could be as follows: in relation to $x^*$, we look for the nearest values to the left and right on the one-dimensional scale, choose the minimum one and continue searching sideways until exactly $K$ neighbours are found.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Constructing a family of Peano curves}\label{MPeano}

@@

\section{Implementation features of the kNN method using multiple space-filling curves}\label{knnme}

The scheme of implementation of the kNN algorithm using multiple Peano curves will look as follows.
For every object $\omega_i$ a set of its preimages on different one-dimensional scales is assigned 
\begin{equation}
\omega_i \rightarrow \{ x_i^0, x_i^1, ..., x_i^L \}
\end{equation} 
where $l, \; 0 \leq l \leq L$, is the number of the one-dimensional interval.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Implementation features of the kNN method using multiple space-filling curves}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Experimental Results}\label{results}

@@

The study was performed using several datasets, both randomly generated for testing and real-world  datasets from the database \url{https://archive.ics.uci.edu/ml/index.php}:
\begin{enumerate}
\item
\textbf{DS-Random-1} -- a set of random data generated using the random module of the Python NumPy library. The data has a uniform distribution and contains values on an interval from 0 to 1.

@@

Two kinds of studies were carried out:
\begin{itemize}
\item The time required to find $K$ nearest neighbours for a test object was measured;
\item The accuracy of recognition of test objects was studied, based on the analysis as to whether their neighbours belong to certain classes.
\end{itemize}

@@

\begin{figure}
\widefigure
\includegraphics[width=1.0\linewidth]{fig3.pdf}
\caption{Running times of the algorithms with $N = 2$\label{fig3}}
\end{figure}

@@

\begin{figure}
\widefigure
\includegraphics[width=1.0\linewidth]{fig4.pdf}
\caption{Running times of the algorithms with $N = 3$\label{fig4}}
\end{figure}

\begin{figure}
\widefigure
\includegraphics[width=1.0\linewidth]{fig5.pdf}
\caption{Running times of the algorithms with $N = 4$\label{fig5}}
\end{figure}

@@

\begin{figure}[h]
\widefigure
\includegraphics[width=0.5\linewidth]{fig6.pdf}
\caption{Comparison of neighbour search times when determining the colour of an image point\label{fig6}}
\end{figure}

@@

The time required to find $K$ nearest neighbours on the CarEvoluation dataset is shown in Fig.~\ref{fig6a}. In this experiment, the number of evolvents used simultaneously was 7. 
\begin{figure}
\widefigure
\includegraphics[width=0.5\linewidth]{fig6a.pdf}
\caption{Comparison of neighbour search times for the CarEvoluation dataset\label{fig6a}}
\end{figure}
The results of the comparison on this dataset also demonstrate that the kNN-ME method is faster compared to the kNN-KD method.

@@

\begin{specialtable}[H] 
  \caption{Percentage of errors in DS-Random-2 recognition}\label{tab:t3}
	\center
	\begin{tabular}{ccccccccc}
\toprule
$K$ & \multicolumn{2}{c}{ $N=2$ } & & \multicolumn{2}{c}{$N=3$} & & \multicolumn{2}{c}{$N=4$} \\
		\noalign{\smallskip} \cline{2-3} \cline{5-6} \cline{8-9} \noalign{\smallskip}
  & kNN-KD & kNN-ME & & kNN-KD & kNN-ME & & kNN-KD & kNN-ME \\
\midrule													
5  & 5 & 5 & & 11 & 12 & & 6 & 8 \\
15 & 4 & 6 & & 10 & 11 & & 10 & 9 \\
25 & 5 & 7 & & 10 & 11 & & 7 & 9 \\
35 & 6 & 6 & & 11 & 13 & & 7 & 10 \\
45 & 7 & 6 & & 10 & 13 & & 7 & 10 \\
\bottomrule
\end{tabular}
\end{specialtable}

@@

The results of measuring the recognition accuracy on the real-world DS-Skin dataset show that on real-world data both methods show the same percentage of misclassification. The specific error percentages obtained with different numbers of neighbours $K$ are shown in Table \ref{tab:t4}.

\begin{specialtable}[H] 
  \caption{Percentage of DS-Skin recognition errors, $N=3$}\label{tab:t4}
	\center
\begin{tabular}{ccc}
\toprule
$K$ & kNN-KD & kNN-ME \\
\midrule													
5 & 0.5 & 0.5 \\
15 & 0.5 & 0.5 \\
25 & 0.5 & 0.5\\
35 & 1 & 1\\
45 & 1 & 1\\
\bottomrule
\end{tabular}
\end{specialtable}

@@

\begin{specialtable}[H] 
  \caption{Average error rate (in percent) for DS-CarEvaluation recognition, $N = 6$}\label{tab:t5}
	\center
\begin{tabular}{ccc}
\toprule
$K$ & kNN-KD & kNN-ME \\
\midrule													
5 & 8 & 11 \\
7 & 9 & 10 \\
9 & 8 & 10 \\
11 & 10 & 11 \\
13 & 9 & 11 \\
\bottomrule
\end{tabular}
\end{specialtable}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}\label{conclusions}

The paper considers a classical machine learning method, the $k$ nearest neighbours (kNN) method, which is widely used in machine learning practice and has been proven to perform very well because of its ``explainability'' (the decision  regarding the choice of an object class can be clearly explained). The results of practical implementation of one of kNN acceleration approaches based on dimensionality reduction using space-filling curves are presented.

@@

\dataavailability{The data analyzed in this study are openly available at\\ \url{https://archive.ics.uci.edu/ml/index.php}.}
